MASCARA: coexpression analysis in data from designed experiments

### 2025-01-20 R/23_02_02_real_data_rice.Rmd
25_01_20
MILP arc consistency KEGG network

```{r, echo = FALSE, include=FALSE}
library(devtools)
library(ggplot2)
library(MetStaT)
library(gASCA)
library(patchwork)
library(readxl)
library(ggfortify)
library(ggrepel)
library(grid)
library(gridExtra)
library(tidyr)
library(tidyverse)
library(reshape2)
library(scales)
library(matrixStats)
library(DESeq2)
library(MASS)
library(glmnet)
library(doParallel)    
library(MUVR) 
library(pheatmap)
library(caret)
library(data.table)
library(plotly)
library(mixOmics)
library(DiscriMiner)
```

@@

ASCA_cands2 <- get_ASCA_cands2(a_c.PCD)
ASCA_cands2 <- ASCA_cands2[which(ASCA_cands2[,2] < 0),]
ASCA_cands2 <- ASCA_cands2[order(ASCA_cands2[,2]),]
ASCA_cands2 <- ASCA_cands2[-which(rownames(ASCA_cands2) %in% pathway),]

@@

>>>>

```{r, echo = FALSE, include=FALSE}
# getwd()
library(devtools)
library(ggplot2)
library(gASCA)
library(patchwork)
library(readxl)
library(ggfortify)
library(ggrepel)
library(grid)
library(gridExtra)
library(tidyr)
library(tidyverse)
library(reshape2)
library(scales)
library(matrixStats)
library(DESeq2)
library(MASS)
library(glmnet)
library(doParallel)    
library(pheatmap)
library(data.table)
library(plotly)
library(pls)
library(pracma)

@@

ASCA_cands2 <- get_ASCA_cands2(a_c.PCD)

ac_top <- ASCA_cands2[1:(nrow(ASCA_cands2)/2),]

ASCA_cands2 <- ASCA_cands2[which(ASCA_cands2[,2] < 0),]
ASCA_cands2 <- ASCA_cands2[order(ASCA_cands2[,2]),]
ASCA_cands2 <- ASCA_cands2[-which(rownames(ASCA_cands2) %in% pathway),]

@@

spikes <- sls[-which(sls %in% baits)]
resids <- res_ASCAplus_nb$residuals[,rownames(ac_top)]

spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
                          resids[,which(colnames(resids) %in% baits)], ncomp = 2)

  #####
  y <- colMeans(spls_res$Yloadings)
  U <- spls_res$projection
  UTP <- (dot(t(U),y)/dot(y,y))

  sPLS_bait_cands <- as.data.frame(UTP[order(UTP, decreasing = TRUE)])

```{r}
library(riceidconverter)

X <- X[,which(colnames(X) %in% colnames(resids))]

symbol <- RiceIDConvert(unique(colnames(X)),'TRANSCRIPTID',toType = 'SYMBOL')
symbol$SYMBOL <- gsub("LOC","osa:",symbol$SYMBOL)
symbol <- symbol[grep("^osa",symbol$SYMBOL),]
# KOIDs <- read.table("https://rest.kegg.jp/link/osa/ko")
path <- read.table("https://rest.kegg.jp/link/osa/pathway")
# module <- read.table("https://rest.kegg.jp/link/osa/module")

colnames(path) <- c("PATH","SYMBOL")
merged <- merge(symbol, path, by = "SYMBOL")

pnum <- as.data.frame(table(merged$PATH))
pnum <- pnum[-which(pnum$Freq < 4),]

merged <- merged[which(merged$PATH %in% pnum$Var1),]

if (!requireNamespace("KEGGREST", quietly = TRUE)) {
  install.packages("BiocManager")
  BiocManager::install("KEGGREST")
}

library(KEGGREST)

kegg_ids <- unique(gsub("path:","",merged$PATH))

# - STUPID FUNCTION BELOW -  ??

# Function to get pathway description for each ID
get_kegg_description <- function(kegg_id) {
  pathway_info <- keggGet(kegg_id)  # Query KEGG
  if (!is.null(pathway_info[[1]])) {
    return(pathway_info[[1]]$NAME)  # Extract the name/description
  } else {
    return(NA)  # Return NA if the ID is invalid
  }
}

# Apply the function to all KEGG IDs
kegg_descriptions <- sapply(kegg_ids, get_kegg_description)

# Combine IDs and descriptions into a data frame
kegg_mapping <- data.frame(
  KEGG_ID = kegg_ids,
  Description = kegg_descriptions,
  stringsAsFactors = FALSE
)

# Print the mapping
print(kegg_mapping)

kegg_DESC <- t(do.call(cbind,kegg_descriptions))

X <- X[,which(colnames(X) %in% merged$TRANSCRIPTID)]       #from ~25k to 4k IDs

cands <- sPLS_bait_cands[which(rownames(sPLS_bait_cands) %in% colnames(X)),, drop = FALSE]
cands <- cbind(rownames(cands), cands)
colnames(cands) <- c("TRANSCRIPTID","UTP")

cands <- merge(cands,merged, by = "TRANSCRIPTID")

cands <- cands[order(cands$UTP, decreasing = TRUE),]

kegg_DESC[gsub("path:","",as.data.frame(table(cands$PATH[c(nrow(cands):(nrow(cands) - 50))]))[,1]),]

kegg_DESC[gsub("path:","",as.data.frame(table(cands$PATH[1:50]))[,1]),]

resids <- resids[,which(colnames(resids) %in% colnames(X))]

res_cor <- cor(resids)
```
```{r}

diag(res_cor) <- NA

threshold <- quantile(res_cor, 0.999, na.rm = TRUE)

inds <- which(res_cor > threshold, arr.ind = TRUE)

cor_vals <- res_cor[inds]

results <- data.frame(
  Row = inds[, 1],
  Column = inds[, 2],
  Correlation = cor_vals
)

if (!is.null(rownames(res_cor))) {
  results$Variable1 <- rownames(res_cor)[results$Row]
  results$Variable2 <- colnames(res_cor)[results$Column]
}

results <- results[results$Row < results$Column, ]
results <- results[order(-results$Correlation), ]

length(unique(c(results$Variable1, results$Variable2)))

res1 <- merge(results,merged, by.x = "Variable1", by.y = "TRANSCRIPTID")
res1 <- merge(res1,merged, by.x = "Variable2", by.y = "TRANSCRIPTID")
res1 <- res1[which(res1$`PATH.x` == res1$`PATH.y`),]

cands <- cands[order(cands$UTP, decreasing = TRUE),]

kegg_DESC[gsub("path:","",as.data.frame(table(res1$`PATH.x`))[,1]),]

same_path <- as.data.frame(table(res1$`PATH.x`))
same_path <- same_path[which(same_path[,2] > 5 & same_path[,2] < 50),]

kegg_DESC[gsub("path:","",same_path[,1]),]

saveRDS(same_path, "path_example.RDS")

### 2025-01-20 R/MASCARA_Tutorial.Rmd
24_02_08_b
24_02_23
Package?
25_01_20
MILP arc consistency KEGG network

```{r, echo=FALSE}
library(devtools)
library(ggplot2)
library(MetStaT)
library(gASCA)
library(patchwork)
library(readxl)
library(ggfortify)
library(ggrepel)
library(grid)
library(gridExtra)
library(tidyr)
library(tidyverse)
library(reshape2)
library(scales)
library(matrixStats)
library(DESeq2)
library(MASS)
library(glmnet)
library(doParallel)    
library(MUVR) 
library(pheatmap)
library(caret)
library(data.table)
library(plotly)
library(mixOmics)
library(DiscriMiner)
library(pracma)
library(matlib)
library(truncnorm)

@@

```{r}
ar <- ASCA(sim_data, m = meta, Baits = baits, spikes = spikes, ncands = ncands, distance_calc = FALSE, Return_Model = FALSE)
resids <- ar[[3]]
```

>>>>

```{r, echo=FALSE}
library(devtools)
library(ggplot2)
library(MetStaT)
library(gASCA)
library(patchwork)
library(readxl)
library(ggfortify)
library(ggrepel)
library(grid)
library(gridExtra)
library(tidyr)
library(tidyverse)
library(reshape2)
library(scales)
library(matrixStats)
library(DESeq2)
library(MASS)
library(glmnet)
library(doParallel)    
# library(MUVR) 
library(pheatmap)
library(caret)
library(data.table)
library(plotly)
# library(mixOmics)
# library(DiscriMiner)
library(pracma)
library(matlib)
library(truncnorm)

@@

```{r}
ar <- ASCA_decompose(d = meta[,1:2], x = sim_data, f = "growth_condition + time + growth_condition:time")
resids <- ar$residuals
```

### 2025-01-20 R/Paper_markdown.Rmd
25_01_20
MILP arc consistency KEGG network

---
title: "MASCARA: coexpression analysis in data from designed experiments"
author: "Fred T.G. White, Anna Heintz-Buschart, Lemeng Dong, Harro J. Bouwmeester, Johan A. Westerhuis, Age K. Smilde"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
    bookdown::word_document2:
      number_sections: false
      toc: false
      classoption:
        twocolumn
bibliography: MASCARA_refs.json
csl: vancouver.csl
keep_md: true
link-citations: yes
header-includes:
  - \usepackage{multicol}
  - \usepackage{amsmath}
  - \usepackage{float}
  - \usepackage{bm}
  - \newcommand{\hideFromPandoc}[1]{#1}
  -  \hideFromPandoc{
       \let\Begin\begin
          \let\End\end
          }

@@

```{r, echo = FALSE, include=FALSE}
library(devtools)
library(ggplot2)
library(MetStaT)
library(gASCA)
library(patchwork)
library(readxl)
library(ggfortify)
library(ggrepel)
library(grid)
library(gridExtra)
library(tidyr)
library(tidyverse)
library(reshape2)
library(scales)
library(matrixStats)
library(DESeq2)
library(MASS)
library(glmnet)
library(doParallel)
library(MUVR)
library(pheatmap)
library(caret)
library(mixOmics)
library(cowplot)
library(reticulate)
library(biomformat)
library(DiscriMiner)
library(viridis)
library(directlabels)
library(ggpmisc)
library(DescTools)
library(data.table)
library(plotly)
library(ggbreak)
library(rhmmer)
library(pracma)
```

@@

From a plant biology point of view, the between-group and the within-group variance can be quite different. However, both may contain valuable biological information. The between-group variance is in principle larger because groups are the result of treatments that are usually selected because they affect the POI. Within-group variance (variance across replicates), if too large, can be an issue in DE analysis, we argue that in co-expression analysis the information present in this variance can be valuable. Variance between replicates is caused by non-controlled variations in plant development and environmental conditions. One replicate plant within a group may be slightly bigger than the others and therefore respond slightly different to a treatment. This could result in slight differences in bait genes between the replicates, but also the genes that we want to identify with a co-expression analysis. There may be many other factors that are inducing this within-group variance, but it is quite likely that if a factor induces a change in the treatment response in one replicate, that this response difference occurs throughout the entire process or pathway in that one replicate. This means that the within-group variance likely contains valuable information and by averaging out this variance, as is done in differential expression analysis, we are losing information. Not making the distinction between types of variance results in *total CoE* associations driven by the differences between treatment groups as opposed to within group correlations. @haiderTranscriptomeAnalysisPhosphate2023 used ranked Pearson correlations to determine (total) coexpression candidates. Due to such a strong effect of the experiment, many genes were affected on top of the desired SL genes. This masked the within-group correlation between the baits and the targets (remaining unknown SL genes). This problematic issue was also highlighted recently by @saccentiWhatCanGo2023 where group sampling assumptions are investigated. Hence, both from a data analysis as well as from a biological perspective it is relevant to make a distinction between the different types of CoE. The focus of this paper will be on developing methods to investigate within-group CoE.

@@

For a successful coexpression analysis it is necessary to have a predefined set of baits that are already known to be involved in the pathway under study. Methods like ranked correlations and partial least squares regression (PLS; @geladiPartialLeastsquaresRegression1986; @woldCollinearityProblemLinear1984) are able to investigate relationships between a set of baits and the other measured genes. In ranked correlations the relationships are quantified at a bivariate level. In PLS regression the relationships can be investigated in a multivariate way, in its simplest form we can see which genes are most predictive of the expression of one bait. This can be extended to a PLS2 model where relationships are investigated between a set of multiple baits and all other genes. These methods do not by default account for the different variance types discussed above. Hence, they are not directly suitable for within-group CoE.

@@

To explore the utility of MASCARA, the method will be compared to ASCA, PLS2 and correlation analysis illustrated using an example RNAseq dataset (from @haiderTranscriptomeAnalysisPhosphate2023). Furthermore, this dataset is used as a guide to several simulation studies to validate the development of a novel approach to coexpression analysis. For these simulations we emulate both different levels of between- and within-group variance. We test the effects of size of between group differences, number of URP differential genes, measurement noise (random unstructured technical variance), structured ambient variance and number of replicates. The following sections seek to investigate which statistical methods are best suited for POI elucidation when a set of baits is already known and illustrate the benefit of MASCARA. Especially in data coming from experiments designed for differential expression analysis.

```{r, echo = FALSE, include=FALSE}
baits <- paste0("X_",c(1989:2000))   #501:510

meta <- cbind.data.frame(rep(c(1,-1), each = 12), rep(c(1:4), each = 3), rep(c(-1:-4,1:4), each = 3))
colnames(meta) <- c("growth_condition","time","interaction")
```
# Materials and Methods

@@

Noise_Tests <- readRDS("Noise_Sim_Coexp_24_01_22.RDS")

names(Noise_Tests) <- c("ASCA", "PLS", "Correlations", "MASCARA")

Noise_Tests <- Noise_Tests[c("ASCA", "PLS", "Correlations","MASCARA")]
Noise_Tests <- lapply(Noise_Tests, narm, keep_col1 = T)

@@

for(i in 2:4){
  res <- Noise_Tests[[i]][12,]
  RES <- cbind.data.frame(RES,t(res))
}

@@

mRES$Noise <- rep(ERs, 4)
mRES$Main <- rep(main_lev, 4)
```

@@

mRES$Size_num <- rep(y, 4)
```

@@

Rep_Tests <- readRDS("Noise_Sim_24_01_22_Replicate_Tests.RDS")

names(Rep_Tests) <- c("ASCA", "PLS","Correlations","MASCARA")

@@

y <- rep(c(0.0004,0.002, 0.01),3)      
x <- rep(25,9)
nse <- rep(c(0.1,0.2,0.3), each = 3)
DS <- c(1:9)

@@

for(i in 2:4){
  res <- Rep_Tests[[i]][12,]
  RES <- cbind.data.frame(RES,t(res))
}

@@

orders <- c(7,8,9,4,5,6,1,2,3)

max <- max(log2(mRES$F1_Score) + 1)
i <- NULL
l <- 1
One_Balance <- list()
for(i in orders){

  One_Balance[[i]] <- ggplot(mRES[which(mRES$nCERs == orders[i]),], aes(x = Replicate_Number, y = log2(F1_Score), colour = Method, shape = Method, fill = Method)) +
  scale_color_viridis(discrete = T)+
    stat_summary() +
    ylim(0,max ) +
  labs(subtitle = paste0("CE = ", main_lev[l], " E = ", nse[l]))+
  theme_dark() +
      theme(axis.title = element_blank())

@@

# tiff("Fig7.tif")
# ps1 <- grid.arrange(ps, left = "log2(GMR)", bottom = "Replicate Number                     ")
# dev.off()

a <- knitr::include_graphics("Fig7.tif")
a

@@

@@

```{r residualdists, fig.cap=c("Distributions of Fisher transformed Pearson correlations (supplementary equation \\@ref(eq:Fisher)) in residuals from the ASCA model on the data from @haiderTranscriptomeAnalysisPhosphate2023. Blue indicates correlations between SL pathway genes, red indicates correlations between SL pathway genes and top 10% DE URP genes (upregulated in P- over P+).")}

@@

@@

a <- knitr::include_graphics("Fig9.tiff")
a

>>>>

---
title: "MASCARA: coexpression analysis in data from designed experiments"
author: "Fred T.G. White, Anna Heintz-Buschart, Lemeng Dong, Harro J. Bouwmeester, Johan A. Westerhuis, Age K. Smilde"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
    bookdown::word_document2:
      number_sections: false
      toc: false
      classoption:
        twocolumn
bibliography: MASCARA_refs.json
csl: vancouver.csl
keep_md: true
link-citations: yes
header-includes:
  - \usepackage{amsmath}
  - \usepackage{float}
  - \usepackage{bm}
  - \newcommand{\hideFromPandoc}[1]{#1}
  -  \hideFromPandoc{
       \let\Begin\begin
          \let\End\end
          }

@@

```{r, echo = FALSE, include=FALSE}
library(devtools)
library(ggplot2)
# library(MetStaT)
library(gASCA)
library(patchwork)
library(readxl)
library(ggfortify)
library(ggrepel)
library(grid)
library(gridExtra)
library(tidyr)
library(tidyverse)
library(reshape2)
library(scales)
library(matrixStats)
library(DESeq2)
library(MASS)
library(glmnet)
library(doParallel)
# library(MUVR)
library(pheatmap)
# library(caret)
# library(mixOmics)
library(cowplot)
library(reticulate)
library(biomformat)
# library(DiscriMiner)
library(viridis)
library(directlabels)
library(ggpmisc)
library(DescTools)
library(data.table)
library(plotly)
# library(ggbreak)
# library(rhmmer)
library(pracma)
library(pls)
```

@@

From a biological point of view, the between-group and the within-group variance can be quite different. However, both may contain valuable biological information. The between-group variance is in principle larger because groups are the result of treatments that are usually selected because they affect the POI. Within-group variance (variance across replicates), if too large, can be an issue in DE analysis, we argue that in co-expression analysis the information present in this variance can be valuable. Variance between replicates is caused by non-controlled variations in plant development and environmental conditions. One replicate plant within a group may be slightly bigger than the others and therefore respond slightly different to a treatment. This could result in slight differences in bait genes between the replicates, but also the genes that we want to identify with a co-expression analysis.
loblolly pine cones @liExtensiveVariationDroughtInduced2021
There may be many other factors that are inducing this within-group variance, but it is quite likely that if a factor induces a change in the treatment response in one replicate, that this response difference occurs throughout the entire process or pathway in that one replicate. This means that the within-group variance likely contains valuable information and by averaging out this variance, as is done in differential expression analysis, we are losing information. Not making the distinction between types of variance results in *total CoE* associations driven by the differences between treatment groups as opposed to within group correlations. @haiderTranscriptomeAnalysisPhosphate2023 used ranked Pearson correlations to determine (total) coexpression candidates. Due to such a strong effect of the experiment, many genes were affected on top of the desired SL genes. This masked the within-group correlation between the baits and the targets (remaining unknown SL genes). This problematic issue was also highlighted recently by @saccentiWhatCanGo2023 where group sampling assumptions are investigated. Hence, both from a data analysis as well as from a biological perspective it is relevant to make a distinction between the different types of CoE. The focus of this paper will be on developing methods to investigate within-group CoE.

@@

For a successful coexpression analysis it is necessary to have a predefined set of baits that are already known to be involved in the pathway under study. Methods like ranked correlations WGCNA  and partial least squares regression (PLS; @geladiPartialLeastsquaresRegression1986; @woldCollinearityProblemLinear1984) are able to investigate relationships between a set of baits and the other measured genes. In ranked correlations the relationships are quantified at a bivariate level. In PLS regression the relationships can be investigated in a multivariate way, in its simplest form we can see which genes are most predictive of the expression of one bait. This can be extended to a PLS2 model where relationships are investigated between a set of multiple baits and all other genes. These methods do not by default account for the different variance types discussed above. Hence, they are not directly suitable for within-group CoE.

@@

To explore the utility of MASCARA, the method will be compared to ASCA, PLS2 and correlation analysis illustrated using an example RNAseq dataset (from @haiderTranscriptomeAnalysisPhosphate2023) as well as an LCMS dataset from @caldanaHighdensityKineticAnalysis2011. Furthermore, this dataset is used as a guide to several simulation studies to validate the development of a novel approach to coexpression analysis. For these simulations we emulate both different levels of between- and within-group variance. We test the effects of size of between group differences, number of URP differential genes, measurement noise (random unstructured technical variance), structured ambient variance and number of replicates. The following sections seek to investigate which statistical methods are best suited for POI elucidation when a set of baits is already known and illustrate the benefit of MASCARA. Especially in data coming from experiments designed for differential expression analysis.

```{r, echo = FALSE, include=FALSE}
baits <- paste0("X_",c(1989:2000))   #501:510
meta <- cbind.data.frame(rep(c(1,-1), each = 12), rep(c(1:4), each = 3), rep(c(-1:-4,1:4), each = 3))
colnames(meta) <- c("growth_condition","time","interaction")
```
# Materials and Methods

@@

# Noise_Tests <- readRDS("Noise_Sim_Coexp_24_01_22.RDS")

Noise_Tests <- readRDS("Noise_Sim_Coexp_24_06_13.RDS")    #Noise_Sim_Coexp_24_05_24.RDS

names(Noise_Tests) <- c("ASCA", "PLS", "Correlations", "MASCARA", "WGCNA")

Noise_Tests <- Noise_Tests[c("ASCA", "PLS", "Correlations", "WGCNA", "MASCARA")]
Noise_Tests <- lapply(Noise_Tests, narm, keep_col1 = T)

@@

for(i in 2:5){
  res <- Noise_Tests[[i]][12,]
  RES <- cbind.data.frame(RES,t(res))
}

@@

mRES$Noise <- rep(ERs, 5)
mRES$Main <- rep(main_lev, 5)
```

@@

mRES$Size_num <- rep(y, 5)
```

@@

# Rep_Tests <- readRDS("Noise_Sim_24_01_22_Replicate_Tests.RDS")

Rep_Tests <- readRDS("Noise_Sim_24_12_08_Replicate_Tests_3.RDS") #Noise_Sim_24_12_04_Replicate_Tests.RDS  #Noise_Sim_24_06_13_Replicate_Tests.RDS Noise_Sim_24_05_24_Replicate_Tests.RDS

names(Rep_Tests) <- c("ASCA", "PLS","Correlations","MASCARA", "WGCNA")

@@

# y <- rep(c(0.0004,0.002, 0.01),3)      
# x <- rep(25,9)
# nse <- rep(c(0.1,0.2,0.3), each = 3)

# y <- rep(c(0.008333333, 0.016666667, 0.025000000),3)      
# x <- rep(30,9)  #100
# nse <- rep(c(0.2,0.4,0.6), each = 3)

m_l <- c(1:5)/10

x <- rep(30,25)  #100  number of Experiment responders (CE) genes  URP

y <- m_l/x[1]     #rep(c(0.008333333, 0.016666667, 0.025000000))
y <- rep(y,length(y))

nse <- rep(m_l, each = length(m_l))  #rep(c(0.1,0.2, 0.3, 0.4, 0.5,0.6), each = length(y))

DS <- c(1:25)

@@

for(i in 2:5){
  res <- Rep_Tests[[i]][12,]
  RES <- cbind.data.frame(RES,t(res))
}

@@

# orders <- c(7,8,9,4,5,6,1,2,3)      # - WARNING -
orders <- c(21,22,23,24,25,
            16,17,18,19,20,
            11,12,13,14,15,
            6,7,8,9,10,
            1,2,3,4,5)

max <- max(log2(mRES$F1_Score) )
i <- NULL
l <- 1
One_Balance <- list()
for(i in orders){

  One_Balance[[i]] <- ggplot(mRES[which(mRES$nCERs == orders[i]),], aes(x = Replicate_Number, y = log2(F1_Score), colour = Method, shape = Method, fill = Method)) +
  scale_color_viridis(discrete = T)+
    stat_summary() +
    ylim(0,max ) +
  labs(subtitle = paste0("CE = ", main_lev[l], " E = ", nse[l]))+
  theme_bw() +
      theme(axis.title = element_blank())

@@

# tiff("Fig7_new1.tif",width = 960*4, height = 960*4,pointsize = 12, res = 300)
# ps1 <- grid.arrange(ps, left = "log2(GMR)", bottom = "Replicate Number                     ")
# dev.off()

a <- knitr::include_graphics("Fig7_new1.tif")
a

getwd()

@@

```{r}
      # F_norms <- c(do.call(c,lapply(res_ASCAplus_nb$decomposition, function(x) norm(x[, c(pathway, non_pathway)], type = "F"))), 
      #            residual = norm(res_ASCAplus_nb$residuals[, c(pathway, non_pathway)], type = "F"))
      #

      F_norms <- c(do.call(c,lapply(res_ASCAplus_nb$decomposition, norm, type = "F")), 
                 residual = norm(res_ASCAplus_nb$residuals, type = "F"))

MASCARA <- function(resids, baits){
  #target projection
  spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
                          resids[,which(colnames(resids) %in% baits)], ncomp = 2)

  #####
  y <- colMeans(spls_res$Yloadings)
  U <- spls_res$projection
  UTP <- (dot(t(U),y)/dot(y,y))

  sPLS_bait_cands <- as.data.frame(UTP[order(UTP, decreasing = TRUE)])

  return(list(sPLS_bait_cands,UTP,spls_res,y))
}

m_res_full <- MASCARA(res_ASCAplus_nb$residuals,baits)
m_rest <- m_res_full[[3]]$scores %*% t(m_res_full[[3]]$loadings)
m_baits <- m_res_full[[3]]$Yscores %*% t(m_res_full[[3]]$Yloadings)

m_comb <- cbind.data.frame(m_rest, m_baits)

deflated_perc <- sum(m_res_full[[3]]$Xvar)/m_res_full[[3]]$Xtotvar

      F_norms <- c(do.call(c,lapply(res_ASCAplus_nb$decomposition, norm, type = "F")), 
                 residual = norm(res_ASCAplus_nb$residuals, type = "F"))

F_norms <- c(F_norms, 
             "CE" = norm(res_ASCAplus_nb$decomposition$growth_condition +
                           res_ASCAplus_nb$decomposition$`growth_condition:time`, type = "F"),
             "modelled" = deflated_perc * F_norms["residual"],
             "noise" = (1-deflated_perc) * F_norms["residual"] )

F_norms[5:7] / sum(F_norms[5:7])

@@

@tajiImportantRolesDrought2002

@zintaDynamicsMetabolicResponses2018

```{r residualdists, fig.cap=c("Distributions of Fisher transformed Pearson correlations (supplementary equation \\@ref(eq:Fisher)) in residuals from the ASCA model on the data from @haiderTranscriptomeAnalysisPhosphate2023. Blue indicates correlations between SL pathway genes, red indicates correlations between SL pathway genes and top 10% DE URP genes (upregulated in P- over P+).")}
library(FNN)

@@

```{r}
GA_ref <- read_xlsx("Rice_Annotation/GA pathway genes-IDs.xlsx")
GA_ref$ID <- gsub("g","t",GA_ref$ID)
GAs <- ref[which(gsub("-.*","",ref$Feature) %in% GA_ref$ID),1]
GAs <- GAs[which(GAs %in% colnames(res_ASCAplus_nb$residuals))]

# pathway <- GAs

# Load necessary libraries
library(reshape2)  # For melt()
library(ggplot2)   # For ggplot()

# Define the function to calculate Fisher-transformed correlation values
cor_val <- function(x){
  cor_res <- cor(x[])
  cor_res <- 0.5 * log((1 + cor_res) / (1 - cor_res))
  return(cor_res)
}

# Calculate correlations for res_ASCAplus_nb$residuals
# test_residuals <- cor_val(res_ASCAplus_nb$residuals[, pathway])

# pathway <- baits

ASCA_cands2_filtered <- ASCA_cands2[!rownames(ASCA_cands2) %in% pathway, ]

# non_pathway <- PSI_ref[,1]
# non_pathway <- non_pathway[which(non_pathway %in% rownames(ASCA_cands2_filtered))]

non_pathway <- rownames(ASCA_cands2_filtered[which(ASCA_cands2_filtered$VIP > quantile(ASCA_cands2_filtered$VIP, 0.9)),])#[1:1253]

test2_residuals <- cor_val(res_ASCAplus_nb$residuals[, c(pathway, non_pathway)])
test2_residuals[upper.tri(test2_residuals, diag = TRUE)] <- NA

mt_residuals <- melt(test2_residuals, na.rm = TRUE)

mt_residuals$group <- "Other"
mt_residuals$group[mt_residuals$Var1 %in% pathway & mt_residuals$Var2 %in% pathway] <- "SL-SL"
mt_residuals$group[mt_residuals$Var1 %in% non_pathway & mt_residuals$Var2 %in% pathway] <- "SL-Other"

mt_residuals <- mt_residuals[mt_residuals$group %in% c("SL-SL", "SL-Other"), ]
colnames(mt_residuals)[3] <- "Correlation"

mt_residuals$Dataset <- "Within"

# test_X <- cor_val(X[, pathway])

test2_X <- cor_val(X[, c(pathway, non_pathway)])
test2_X[upper.tri(test2_X, diag = TRUE)] <- NA

mt_X <- melt(test2_X, na.rm = TRUE)

mt_X$group <- "Other"
mt_X$group[mt_X$Var1 %in% pathway & mt_X$Var2 %in% pathway] <- "SL-SL"
mt_X$group[mt_X$Var1 %in% non_pathway & mt_X$Var2 %in% pathway] <- "SL-Other"

mt_X <- mt_X[mt_X$group %in% c("SL-SL", "SL-Other"), ]
colnames(mt_X)[3] <- "Correlation"

mt_X$Dataset <- "Total"

#deflated residuals:
# MASCARA <- function(resids, baits){
#   #target projection
#   spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
#                           resids[,which(colnames(resids) %in% baits)], ncomp = 3)
#   
#   #####
#   y <- colMeans(spls_res$Yloadings)
#   U <- spls_res$projection
#   UTP <- (dot(t(U),y)/dot(y,y))
#   
#   sPLS_bait_cands <- as.data.frame(UTP[order(UTP, decreasing = TRUE)])
#   
#   return(list(sPLS_bait_cands,UTP,spls_res,y))
# }

# m_res <- MASCARA(res_ASCAplus_nb$residuals[, c(pathway, non_pathway)],baits)[[3]]
# 
# m_rest <- m_res$scores %*% t(m_res$loadings)
# m_baits <- m_res$Yscores %*% t(m_res$Yloadings)
# 
# m_comb <- cbind.data.frame(m_rest, m_baits)
# 
# test_m <- cor_val(m_comb[, c(pathway, non_pathway)])
# test_m[upper.tri(test_m, diag = TRUE)] <- NA
# 
# mt_m <- melt(test_m, na.rm = TRUE)
# 
# mt_m$group <- "Other"
# mt_m$group[mt_m$Var1 %in% pathway & mt_m$Var2 %in% pathway] <- "SL-SL"
# mt_m$group[mt_m$Var1 %in% non_pathway & mt_m$Var2 %in% pathway] <- "SL-Other"
# 
# mt_m <- mt_m[mt_m$group %in% c("SL-SL", "SL-Other"), ]
# colnames(mt_m)[3] <- "Correlation"
# 
# mt_m$Dataset <- "Deflated"
#

library(ggpubr)

mt_combined <- rbind(mt_residuals, mt_X)   #   , mt_m

mt_combined$Group_Dataset <- paste(mt_combined$group, mt_combined$Dataset, sep = "_")

mt_combined$Dataset <- factor(mt_combined$Dataset, levels = c("Total", "Within", "Deflated"))

dens1 <- ggplot(mt_combined, aes(x = group, y = Correlation, fill = group)) +
  geom_violin() +
  theme_bw() +
  labs(fill = "Group") +
  ggtitle("Boxplot of Fisher-Transformed Correlations by Variance Partition") +
  xlab("Group") +
  ylab("Correlation") +
  facet_wrap(~ Dataset) +
  stat_compare_means() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none"  # Remove legend if 'fill' is the same as 'x'
  )

print(dens1)

dens <- ggplot(mt_combined, aes(Correlation, fill = group)) + geom_histogram(bins = 140) + theme_bw()  + facet_wrap(~ Dataset)

dens

# 
# xkl <- KL.divergence(mt_combined$Correlation[which(mt_combined$Group_Dataset == "SL-Other_Total")], mt_combined$Correlation[which(mt_combined$Group_Dataset == "SL-SL_Total")])
# 
# rkl <- KL.divergence(mt_combined$Correlation[which(mt_combined$Group_Dataset == "SL-Other_Within")], mt_combined$Correlation[which(mt_combined$Group_Dataset == "SL-SL_Within")])
# 
# plot(xkl, rkl)
# abline(a=0,b=1)
```

@@

a <- knitr::include_graphics("Fig9.tif")
a
